A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2019-06-24 paper/paper.Rnw
minor edits

<<setup, include = F>>=
knitr::opts_chunk$set(echo = F, dev = "pdf", dpi = 1000, message = F, warning = F, fig.align = "center")

@@

In this paper, we discuss an alternate approach \mt{using the base of a pre-trained convolutional neural network and training a} an additional model layer to transfer the feature vector output into a vector of probabilities representing the detection of geometric elements in shoe tread images. Working within the feature space used by forensic examiners allows us to augment human-identified features with model output, assessing similarity of different shoe images on a feature set that is explicitly relevant to the domain.
In addition, much of the existing software to assess geometric similarity, such as SoleMate\footnote{\url{http://www.fosterfreeman.com/trace-evidence/357-sicar-6-solemate-2.html}}, requires examiners to identify and label features manually; automating classification within the currently used feature space could reduce or eliminate the need for such manual classification.

@@

\section{Materials and Methods}
\fix{We may want to mention some of the methods in the introduction for continuity sake...}
\mt{I'm not sure if adding a sentence to the above paragraph accomplished this goal. Do you think I need to write a new transition paragraph instead? I tried doing that (see commented out stuff), but wasn't sure how to avoid redundancy with the methods section...}

@@

\paragraph{Transfer Learning}
Neural networks are composed of sets of layers; the early layers, which are called the model base, contain feature detectors, and the final layers make up the classifier, or model head, which connects meaningful features to classification labels (see \autoref{fig:VGG16-structure} for an example of the structure of a CNN). \mt{Each of these layers is comprised of sets of weights and connections, or parameters, that are optimized during training.} A relatively simple convolutional neural network which processes image data may contain more than 14 million parameters in the model base and an additional 120 million parameters in the model head; to train a model of this nature requires millions of labeled images and a significant amount of computational power. Assembling sufficient data to train a network from scratch is a gargantuan task; the process of \emph{transfer learning} is a natural solution to this common problem. Transfer learning leverages the modularity of neural networks, that is, that the model base that detects features can be separated from the classifier which produces predictions. Transfer learning approaches use the base of a model trained on more general image data and train a new model head or classifier using a smaller set of domain-specific labeled images \mt{\citep{oquabLearningTransferringMidlevel2014}, allowing CNNs to be applied to smaller datasets of several thousand images, reduces the amount of computational time required to fit the model, and provides boosted performance compared to training a new model from scratch \citep{yosinskiHowTransferableAre2014}. It has been successfully applied in automatic classification of medical images \citep{shinDeepConvolutionalNeural2016} as well as in various applications of shoe forensics \citep{kongCrossDomainForensicShoeprint2017,kongCrossDomainImageMatching2019a,zhangAdaptingConvolutionalNeural2017}.}

Many convolutional neural networks have been trained on a subset of images from an online image database called ImageNet \citep{dengImageNetLargeScaleHierarchical}. In particular, there is a standard set of images from ImageNet, consisting of 1.2 million images \mt{that span} 1,000 catetories, that is commonly used in computer vision tasks\mt{\sout{; these images provide both a wide variety of images for training complex image-classification models and a standardized way to compare performance across different models}}. CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. In the next section, we will introduce the pre-trained network VGG16, pictured in \autoref{fig:VGG16-structure}, and examine a selection of its filters (shown in \autoref{vgg16layers}) that demonstrate the levels of detail in the initial layers of a model trained on generic data.

@@

\paragraph{VGG16}
Developed by Oxford's Visual Graphics Group, VGG16 is a CNN trained on ImageNet
\citep{krizhevskyImageNetClassificationDeep2012}. In contrast to other popular \mt{pre-trained} networks, like ResNet, VGG has a relatively simple structure that provides easier training and interpretability with very little sacrificed accuracy. The simplicity of this structure provides the ability to peer into the inner workings of the network for diagnostic purposes, providing a distinct advantage over more complicated network structures with slightly higher accuracy ratings. VGG16 is a common choice for transfer learning because of this structural simplicity; it has been used for detection of text in natural images \citep{zhongDeepTextUnifiedFramework2016}, medical imaging classification \citep{oquabLearningTransferringMidlevel2014}, classification of weld defects \citep{liuWeldDefectImages2018}, and many other domain-specific image recognition tasks that are more specific than the ImageNet data on which it was trained. \autoref{fig:VGG16-structure} shows the architecture of VGG16; with transfer learning, the VGG16 model head is replaced with a trained model head specific to recognition of shoeprint class characteristics.

\mt{The base of VGG16 is made up of five convolutional blocks, each of which contain between 128 and 1,536 filters.} A filter is a set of numerical weights that, when applied to an image, quantifies the presence of the specific shapes and/or colors that the filter is trained to detect. \autoref{fig:vgg16layers} shows a selection of filters from VGG16, with each row corresponding to one of the five convolutional blocks. Filters in earlier convolutional blocks detect simple features, such as colors, lines, corners, and blobs, while later filters detect more complex combinations of features. This aggregation of features through successive convolutional layers mirrors the process of complex feature detection in human vision. Any complex feature can be disassembled into a set of simple features; similarly, the successive sets of layers detect increasingly detailed features.

@@

We have proposed several hypotheses to explain why the model's predictions are not entirely consistent with manual labels on the images. We can test these hypotheses by examining what the model ``sees" when making its predictions, using heatmaps, or class activation maps (CAMs). \fix{Heatmaps or Class Activation Maps} are a visual diagnostic tool that highlights the areas of an image that are most significant to classification for a given class \citep{chollet_allaire_2018}. Using these graphics, we can identify which locations in an image contribute most to the model's predictions. In \fix{the figures in this subsection,} heatmaps are shown for a selection of three classes, which are either contained in the image or have high output probabilities, alongside the original image. In these heatmaps, we can confirm that multiple classes have common features, resulting in misclassifications.

@@

# calc_heatmap(imgs[which(grepl("Ugg-grey", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("text", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("polygon", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("corks", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("chevron-circle", imgs))], loaded_model) %>%
#   #prune_heatmap(c("chevron", "quad", "circle")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("triangle", imgs))], loaded_model) %>%
#   #prune_heatmap(c("triangle", "quad", "chevron")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("star", imgs))], loaded_model) %>%
#   #prune_heatmap(c("star", "bowtie", "polygon")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("other", imgs))], loaded_model) %>%
#   #prune_heatmap(c("star", "chevron", "polygon")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)

@@

<<current-heatmaps, out.width = ".95\\textwidth", fig.width = 8, fig.height = 6, fig.cap = "Correct or reasonably incorrect heatmaps", echo = F, message = F, warning = F, fig.pos = "!p">>=

@@

knitr::include_graphics(heatmaps[grepl("Ugg-grey", heatmaps)])
knitr::include_graphics(heatmaps[grepl("text", heatmaps)])
knitr::include_graphics(heatmaps[grepl("polygon", heatmaps)])
knitr::include_graphics(heatmaps[grepl("corks", heatmaps)])

@@

% Discussion of weaknesses
\fix{Although model predictions are overall quite good, there are a number of \fix{issues}. Firstly, the model has substantial difficulty predicting images with low contrast because image features are obscured; exploring methods of color correction, such as histogram normalization, may prove useful for eliminating the effect of contrast on predictions.

A larger difficulty is the inherent overlap between features of common geometric shapes, and high variability of those shapes on outsoles. Overlapping features between shapes makes it more difficult for the model to distinguish between geometrically similar shapes, such as chevrons and triangles. When the physical shapes present within a category are highly variable, this further widens the sets of features associated with individual categories, leading to even more model confusion. These issues of overlap and variabailty are at least in part due to the features used by practitioners, and partially due to our generalization/operational definition.}

\mt{What is the balance between comprehensive and redundant? I don't think I've found it... I'll try again when not exhausted}

\mt{Final paragraph will suggest redefining shapes (yuck) and cutting off the final conv block to see whether those features are easier for the model to sort between categories.}

>>>>

<<setup, include = F>>=
knitr::opts_chunk$set(echo = F, dev = "pdf", dpi = 1000, message = F, warning = F, fig.align = "center", fig.path = "figure/")

@@

In this paper, we discuss an alternate approach that uses an additional model layer to transfer the feature vector output into a vector of probabilities representing the detection of geometric elements in shoe tread images. Working within the feature space used by forensic examiners allows us to augment human-identified features with model output, assessing similarity of different shoe images on a feature set that is explicitly relevant to the domain.
In addition, much of the existing software to assess geometric similarity, such as SoleMate\footnote{\url{http://www.fosterfreeman.com/trace-evidence/357-sicar-6-solemate-2.html}}, requires examiners to identify and label features manually; automating classification within the currently used feature space could reduce or eliminate the need for such manual classification.

@@

\section{Materials and Methods}
\fix{We may want to mention some of the methods in the introduction for continuity sake...}
\mt{I'm not sure if adding a sentence to the above paragraph accomplished this goal. Do you think I need to write a new transition paragraph instead? I tried doing that (see commented out stuff), but wasn't sure how to avoid redundancy with the methods section...} \svp{I was talking about the methods above (e.g. fractal decomp, scale invariant feature recognition, etc., not methods we used)}

@@

\paragraph{Transfer Learning}
Neural networks are composed of sets of layers; the early layers, which are called the model base, contain feature detectors, and the final layers make up the classifier, or model head, which connects meaningful features to classification labels (see \autoref{fig:VGG16-structure} for an example of the structure of a CNN). Each of these layers is \svp{composed} of sets of weights and \svp{connections that} are optimized during training. A relatively simple convolutional neural network which processes image data may contain more than 14 million parameters in the model base and an additional 120 million parameters in the model head; to train a model of this nature requires millions of labeled images and a significant amount of computational power. Assembling sufficient data to train a network from scratch is a gargantuan task; the process of \emph{transfer learning} is a natural solution to this common problem. Transfer learning leverages the modularity of neural networks, that is, that the model base that detects features can be separated from the classifier which produces predictions. Transfer learning approaches use the base of a model trained on more general image data and train a new model head or classifier using a smaller set of domain-specific labeled images\citep{oquabLearningTransferringMidlevel2014}. \svp{This allows} CNNs to be applied to smaller datasets of several thousand images, reduces the amount of computational time required to fit the model, and provides boosted performance compared to training a new model from scratch \citep{yosinskiHowTransferableAre2014}. Transfer learning has been successfully applied in automatic classification of medical images \citep{shinDeepConvolutionalNeural2016} as well as in various applications of shoe forensics \citep{kongCrossDomainForensicShoeprint2017,kongCrossDomainImageMatching2019a,zhangAdaptingConvolutionalNeural2017}.

Many convolutional neural networks have been trained on a subset of images from an online image database called ImageNet \citep{dengImageNetLargeScaleHierarchical}. In particular, there is a standard set of images from ImageNet, consisting of 1.2 million images \mt{that span} 1,000 catetories, that is commonly used in computer vision tasks. CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. In the next section, we will introduce the pre-trained network VGG16, pictured in \autoref{fig:VGG16-structure}, and examine a selection of its filters (shown in \autoref{fig:vgg16layers}) that demonstrate the levels of detail in the initial layers of a model trained on generic data.

@@

\paragraph{VGG16}
Developed by Oxford's Visual Graphics Group, VGG16 is a CNN trained on ImageNet
\citep{krizhevskyImageNetClassificationDeep2012}. In contrast to other popular pre-trained networks, \svp{such as AlexNet or ResNet}, VGG has a relatively simple structure that provides easier training and interpretability with very little sacrificed accuracy. The simplicity of this structure provides the ability to peer into the inner workings of the network for diagnostic purposes, providing a distinct advantage over more complicated network structures with slightly higher accuracy ratings. VGG16 is a common choice for transfer learning because of this structural simplicity; it has been used for detection of text in natural images \citep{zhongDeepTextUnifiedFramework2016}, medical imaging classification \citep{oquabLearningTransferringMidlevel2014}, classification of weld defects \citep{liuWeldDefectImages2018}, and many other domain-specific image recognition tasks that are more specific than the ImageNet data on which it was trained. \autoref{fig:VGG16-structure} shows the architecture of VGG16; with transfer learning, the VGG16 model head is replaced with a model head \svp{trained specifically to recognize} shoeprint class characteristics.

The base of VGG16 is made up of five convolutional blocks, each of which contain between 128 and 1,536 filters. A filter is a set of numerical weights that, when applied to an image, quantifies the presence of the specific shapes and/or colors that the filter is trained to detect. \autoref{fig:vgg16layers} shows a selection of \svp{images which maximally activate specific} filters from VGG16, % Note - they aren't actually the filters themselves, these are generated images which maximize activation of the filter in question.
with each row corresponding to one of the five convolutional blocks. Filters in earlier convolutional blocks detect simple features, such as colors, lines, corners, and blobs, while later filters detect more complex combinations of features. This aggregation of features through successive convolutional layers mirrors the process of complex feature detection in human vision. Any complex feature can be disassembled into a set of simple features; similarly, the successive sets of layers detect increasingly detailed features.

@@

We have proposed several hypotheses to explain why the model's predictions are not entirely consistent with manual labels on the images. We can test these hypotheses by examining what the model ``sees" when making its predictions, using heatmaps, or class activation maps (CAMs). Heatmaps, \svp{also known as} Class Activation Maps\svp{,} are a visual diagnostic tool that highlights the areas of an image that are most significant to classification for a given class \citep{chollet_allaire_2018}. Using these graphics, we can identify which locations in an image contribute most to the model's predictions. In \fix{the figures in this subsection,} heatmaps are shown for a selection of three classes, which are either contained in the image or have high output probabilities, alongside the original image. In these heatmaps, we can confirm that multiple classes have common features, resulting in misclassifications.

@@

# calc_heatmap(imgs[which(grepl("Ugg-grey", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("text", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("polygon_recolor", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("corks", imgs))], loaded_model) %>%
#   prune_heatmap(top_n = 3, sort = "value") %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F,
#                    eer = eers)
#
# calc_heatmap(imgs[which(grepl("chevron-circle", imgs))], loaded_model) %>%
#   #prune_heatmap(c("chevron", "quad", "circle")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("triangle", imgs))], loaded_model) %>%
#   #prune_heatmap(c("triangle", "quad", "chevron")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("star", imgs))], loaded_model) %>%
#   #prune_heatmap(c("star", "bowtie", "polygon")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)
#
# calc_heatmap(imgs[which(grepl("other", imgs))], loaded_model) %>%
#   #prune_heatmap(c("star", "chevron", "polygon")) %>%
#   create_composite(save_file = T, outdir = h_imgdir, color_by_label = F)

@@

<<current-heatmaps, out.width = ".95\\textwidth", fig.width = 8, fig.height = 6, fig.cap = c("A section of an Ugg shoe containing the logo. The model correctly identifies triangles and text (and references the correct portion of the image in both cases). The model identifies a circle as well; while it could be argued that the triangles form an implicit circle, the heatmap demonstrates that the circular portion of the image is the 'G', which would not be a closed figure labeled as a circle. The model's prediction of the presence of a circle is understandable, but incorrect.", "In this image of a Seychelles shoe, the model identifies text and circle. The prediction of text is correct; the prediction of circle is not - there are very round shapes, but they do not form a closed figure that could be identified as a circle.", "polygon", "corks"), echo = F, message = F, warning = F, fig.pos = "!p">>=

@@

knitr::include_graphics(heatmaps[grepl("Ugg-grey", heatmaps)])
knitr::include_graphics(heatmaps[grepl("text", heatmaps)])
knitr::include_graphics(heatmaps[grepl("polygon_recolor", heatmaps)])
knitr::include_graphics(heatmaps[grepl("corks", heatmaps)])

@@

% Discussion of weaknesses
\fix{Although model predictions are} \svp{generally accurate}, there are a number of \svp{systematic misclassifications}. The model has substantial difficulty predicting images with low contrast because image features are obscured. Exploring methods of color correction, such as histogram equalization, may prove useful for eliminating the effect of contrast on predictions.

\svp{Another source of systematic misclassifications stems from the} inherent overlap between features of common geometric shapes, \svp{in combination with} high variability of those shapes on outsoles. Overlapping features between shapes makes it more difficult for the model to distinguish between geometrically similar shapes, such as chevrons and triangles. When the physical shapes present within a category are highly variable, this further widens the sets of features associated with individual categories, leading to even more model confusion. These issues of overlap and variability are at least in part due to \svp{our decision to use} the features used by practitioners \svp{directly}; \svp{some of this may also result from our operational definitions of various shapes, which did not take into account the model's capabilities and initial training}.

% \mt{What is the balance between comprehensive and redundant? I don't think I've found it... I'll try again when not exhausted} I'll let you know if I ever find it, but this looks pretty great - I've made a few tiny tweaks to phrasing, but the content is exactly what I was hoping for :)

\mt{Final paragraph will suggest redefining shapes (yuck) and cutting off the final conv block to see whether those features are easier for the model to sort between categories.} \svp{Sounds good!}
